\documentstyle[apaua]{article}
\pagestyle{empty}

\oddsidemargin=0in
\evensidemargin=0in
\topmargin=0.5in
\textheight=9in
\textwidth=6.5in 
\footheight=0in
\columnsep=0.25in
\headsep=0in
\headheight=0in

\def\btt#1{\bf{\tt #1}}

\begin{document}
\bibliographystyle{apaua}

\begin{titlepage}
\begin{center}
\vspace*{2.35in}
SGA-C: A C-language Implementation \\ 
of a Simple Genetic Algorithm
\ \\
by\\
\ \\
Robert E. Smith\\
The University of Alabama\\
\ \\ 
David E. Goldberg\\
The University of Illinois\\
and\\
Jeff A. Earickson\\
Alabama Supercomputer Network\\
\vspace{0.7in}
TCGA Report No. 91002\\
\today \\
\vspace{2.2in}
The Clearinghouse for Genetic Algorithms\\
The University of Alabama\\
Department of Engineering Mechanics\\
Tuscaloosa, AL 35487
\end{center}
\end{titlepage}


\title{SGA-C: A C-language Implementation of a\\ 
Simple Genetic Algorithm}
\author{{\bf Robert E. Smith}\\
The University of Alabama\\
Department of Engineering Mechanics\\
Tuscaloosa, Alabama 35405 
\and {\bf David E. Goldberg}\\
The University of Illinois\\
Department of General Engineering\\
Urbana, Illinois 61801
\and {\bf Jeff A. Earickson}\\
Alabama Supercomputer Network\\
The Boeing Company\\
Huntsville, Alabama 35806}
\maketitle

\section{Introduction}
SGA-C\footnote{{\bf Disclaimer:} SGA-C is distributed under the terms described
in the file {\btt{NOWARRANTY}}. These terms are taken from the GNU General Public
License.
This means that SGA-C has no warranty implied or given, and that the
authors assume no liability for damage resulting from its use or misuse.} 
is a C-language translation and extension of the original
Pascal SGA code presented by Goldberg \citeyear{Goldberg:89e}. 
It has some additional features, but its
operation is 
essentially the same as that of the original, Pascal version. 
This report is included as a concise introduction to the SGA-C distribution.
It is presented with the assumptions that the reader has a general understanding
of Goldberg's original Pascal SGA code, and a good working knowledge of the
C programming language.
The report begins with an outline of the files included in the SGA-C
distribution, and the routines they contain. 
The outline is followed by a discussion of significant features of SGA-C
that differ from those of the Pascal version.
The report concludes with a
discussion of routines that must be altered to
implement one's own application in SGA-C.


\section{Files Distributed with SGA-C}
\label{files}
The following is an outline of the files distributed 
with SGA-C, the routines contained in those files, and the
{\btt{include}} structure of the SGA-C distribution.
\begin{description}
\item[{\btt{sga.h}}] contains declarations of global variables and structures 
for SGA-C.  This file is included by {\btt{main()}}.
Both {\btt{sga.h}} and {\btt{external.h}} have two
{\btt{defines}} set at the top of the files; {\btt{LINELENGTH}}, which determines the
column width of printed output, and {\btt{BITS\_PER\_BYTE}}, which specifies the number
of bits per byte on the machine hardware.  {\btt{LINELENGTH}} can be set to any
desired positive value, but {\btt{BITS\_PER\_BYTE}} must be set to the correct value for your
hardware.
\item[{\btt{external.h}}] contains external declarations for inclusion
in all source code files except {\btt{main()}}.  The {\bf extern} declarations in {\btt{external.h}}
should match the declarations 
in {\btt{sga.h}}.
\item[{\tt main.c}] contains the main SGA program loop, 
{\btt{main()}}.
\item[{\btt{generate.c}}] contains {\btt{generation()}}, a routine which generates and 
evaluates a new GA population.  
\item[{\btt{initial.c}}] contains routines that are called at the beginning 
of a GA run.
\begin{description}
\item[{\btt{initialize()}}] is the central initialization routine called by 
{\btt{main()}}.
\item[{\btt{initdata()}}] is a routine to prompt the user for SGA parameters.
\item[{\btt{initpop()}}] is a routine that generates a random population. Currently, 
SGA-C includes no facility for using seeded populations.
\item[{\btt{initreport()}}] is a routine that prints a report after initialization 
and before the first GA cycle.
\end{description}
\item[{\btt{memory.c}}] contains routines for dynamic memory management.  
\begin{description}
\item[{\btt{initmalloc()}}] is a routine that dynamically allocates space for the GA 
population and other necessary data structures.
\item[{\btt{freeall()}}] frees all memory allocated by {\btt{initmalloc()}}.
\item[{\btt{nomemory()}}] prints out a warning statement
when a call to {\btt{malloc()}} fails.
\end{description}
\item[{\btt{operators.c}}]  contains the routines for genetic operators.
\begin{description}
\item[{\btt{crossover()}}] performs single-point crossover on two mates, producing
two children.
\item[{\btt{mutation()}}] performs a point mutation.
\end{description}
\item[{\btt{random.c}}] contains random number utility programs, including:
\begin{description}
\item[{\btt{randomperc()}}] returns a single, uniformly-distributed, real, 
pseudo-random number between 0 and 1. 
This routine uses the subtractive method specified by Knuth \citeyear{Knuth:81}.
\item[{\btt{rnd(low,high)}}] returns an uniformly-distributed integer between 
{\btt{low}} and {\btt{high}}.
\item[{\btt{rndreal(low,high)}}] returns an uniformly-distributed floating point number 
between  {\btt{low}} and {\btt{high}}.
\item[{\btt{flip(p)}}] flips a biased coin, returning 1 with probability {\btt{p}}, 
and 0 with probability {\btt{1-p}}.
\item[{\btt{advance\_random()}}] generates a new batch of 55 random numbers.
\item[{\btt{randomize()}}] asks the user for a random number seed.
\item[{\btt{warmup\_random()}}] primes the random number generator.
\item[{\btt{noise(mu, sigma)}}] generates a normal random variable 
with mean mu and standard
deviation sigma.  This routine is not currently used in SGA-C, and is only included as a general utility.
\item[{\btt{randomnormaldeviate()}}] is a utility routine used by noise.
It computes a standard normal random variable.
\item[{\btt{initrandomnormaldeviate()}}] initialization routine for
{\btt{randomnormaldeviate()}}.
\end{description}
\item[{\btt{report.c}}] contains routines used to print a report from each cycle of SGA-C's operation.
\begin{description}
\item[{\btt{report()}}] controls overall reporting.
\item[{\btt{writepop()}}] writes out the population at every generation.
\item[{\btt{writechrom()}}] writes out the chromosome as a string of ones and zeroes.
In the current implementation, the most significant bit is the {\it rightmost} bit.
\end{description}
\item Three selection routines are included with the SGA-C distribution:
\begin{description}
\item[{\btt{rselect.c}}] contains routines for roulette-wheel selection.  
\item[{\btt{srselect.c}}] contains the routines for stochastic-remainder selection 
\cite{Booker:82}.
\item[{\btt{tselect.c}}] contains the routines for tournament selection 
\cite{Brindle:81a}.  Tournaments of any size up to the population size can be held
with this implementation\footnote{The tournament selection routine
included with the distribution was written by Hillol Kargupta, of the University
of Alabama.}.
\end{description}
For modularity, each selection method 
is made available as a compile time option.
Edit {\btt{Makefile}} to choose a selection method. Each of the three
selection files 
contains the routines
{\btt{select\_memory}} and {\btt{select\_free}} (called by {\btt{initmalloc}} and {\btt{freeall}}, respectively), which perform
any necessary auxiliary memory handling,
and the routines {\btt{preselect()}} and {\btt{select()}}, 
which implement the particular selection method.
\item[{\btt{stats.c}}] contains the routine {\btt{statistics()}}, which calculates 
populations statistics for each generation.
\item[{\btt{utility.c}}] contains various utility routines. Of particular interest 
is the routine {\btt{ithruj2int()}}, which returns bits $i$ through $j$ of a 
chromosome interpreted as an {\btt{int}}. 
\item[{\btt{app.c}}] contains application dependent routines. 
Unless you need to change the basic operation of the GA itself, 
you should only have to alter this file
Further instructions for altering the SGA application are 
included in section~\ref{app}.
\begin{description}
\item[{\btt{application()}}] should contain any application-specific computations
needed before each GA cycle.  It is called by {\btt{main()}}.
\item[{\btt{app\_data()}}] should ask for and read in any application-specific 
information.  This routine is 
called by {\btt{init\_data()}}.
\item[{\btt{app\_malloc()}}] should perform any application-specific calls to 
{\btt{malloc()}} to dynamically allocate memory.  This routine is 
called by {\btt{initmalloc()}}.
\item[{\btt{app\_free()}}]  should perform any application-specific calls to 
{\btt{free()}}, for release of dynamically allocated memory.  
This routine is called by {\btt{freeall()}}.
\item[{\btt{app\_init()}}] should perform any application-specific initialization
needed.  It is called by {\btt{initialize()}}.
\item[{\btt{app\_initreport()}}] should print out an application-specific initial
report before the start of generation cycles.  This routine is 
called by {\btt{initialize()}}.
\item[{\btt{app\_report()}}]  should print out any application-specific  
output 
after each GA cycle.  It is called by {\btt{report()}}.
\item[{\btt{app\_stats()}}]  should perform any application-specific statistical
calculations.  It is called by {\btt{statistics()}}.
\item[{\btt{objfunc(critter)}}]  The objective function for the specific application.
The variable {\btt{critter}} is a pointer to an {\btt{individual}} 
(a GA population member), to which this routine must assign a fitness.
This routine is called by {\btt{generation()}}.
\end{description}
\item[{\btt{Makefile}}] is a UNIX makefile for SGA-C.  
\end{description}

\section{New Features of SGA-C}
SGA-C has several features that differ from those of the Pascal version.
One is the ability to name the input and output files on the command line, i.e.
{\btt{sga my.input my.output}}.  If either of these files is not named on the command line, 
SGA-C assumes 
{\btt{stdin}} and {\btt{stdout}}, respectively.  
Another new feature of SGA-C is its method of representing
chromosomes in memory.
SGA-C stores its chromosomes in bit strings at the machine level. 
Input-output and chromosome storage in SGA-C are discussed in the following sections.

\subsection{Input-Output}
SGA-C allows for multiple GA runs.
When the program is executed, the user is first prompted
for the number of GA runs to be
performed.  After this, the quantity of input needed depends on 
the selection routine chosen at compile-time, and any
application-specific information required.  
When compiled
with roulette wheel selection, 
the input requested from the user is as follows:
\begin{itemize}
\item The number of GA runs to be performed ({\btt{int}}).
\item The population size ({\btt{int}}).
\item The chromosome length ({\btt{int}}).
\item Print the chromosome strings each generation ({\btt{y/n}})?
\item The maximum number of generations for the run ({\btt{int}}).
\item The probability of crossover ({\btt{float}}).
\item The probability of mutation ({\btt{float}}).
\item Application-specific input, if any.
\item The seed for the random number generator ({\btt{float}}).
\end{itemize}

\subsection{Chromosome Representation and Memory Utilization}
\label{memstuff}
SGA-C uses a machine level representation of bit strings
to increase efficiency.
This allows crossover and mutation to be implemented 
as binary masking operations (see {\btt{operators.c}}).
Every chromosome (as well as the population arrays and some
auxiliary memory space) are allocated dynamically at run
time. The dynamic memory allocation scheme allocates
a sufficient number of unsigned integers for each population member
to store bits for the user-specified chromosome length. Because
of this feature, it is extremely important that {\btt{BITS\_PER\_BYTE}}
be properly set (in
{\btt{sga.h}} and {\btt{external.h}})
for your machine's hardware and C compiler.

\section{Implementing Application Specific Routines}
\label{app}
To implement a specific application, you should only have to change the file
{\btt{app.c}}.
Section~\ref{files} describes the routines in {\btt{app.c}} in detail.
If you use additional variables for your specific problem, the easiest method
of making them available to other program units is to declare them in 
{\btt{sga.h}} and {\btt{external.h}}.  However, take care that you do not
redeclare existing variables.

Two example applications files are included in the SGA-C distribution.  The
file {\btt{app1.c}} performs the simple example problem 
included with the Pascal version; 
finding the maximum of $x^{10}$, where $x$
is an integer interpretation of a chromosome.  
A slightly more complex 
application 
is include in {\btt{app2.c}}.
This application illustrates two features that have been added
to SGA-C. The first of these is the {\bf ithruj2int} function,
which converts bits $i$ through $j$ in a chromosome to an integer.
The second new feature is the {\bf utility} pointer that is associated with each population member.
The example application interprets each chromosome as a set 
of concatenated integers in binary form. The lengths of these integer fields is 
determined by the user-specified value of {\bf field\_size}, which is 
read in by the function
{\btt{app\_data()}}.  The field size must be less than the smallest of
the chromosome
length and the length of an unsigned integer.  
An integer array for storing the interpreted form of each chromosome
is dynamically allocated and assigned to the chromosome's {\bf utility} pointer
in {\btt{app\_malloc()}}.
The {\btt{ithruj2int}} routine (see {\btt{utility.c}}) is used to translate 
each chromosome into its associated vector.
The fitness for each chromosome is simply the sum of the squares of these 
integers. This example application will function for any chromosome length. 

\section{Final Comments}
SGA-C is intended to be a simple
program for first-time GA experimentation. It is 
not intended to be 
definitive in terms of its efficiency 
or the grace of its implementation. The
authors are interested in the comments, criticisms, and bug reports
from SGA-C users, so that the code can be refined for
easier use in subsequent versions.
Please email your comments to {\bf rob@comec4.mh.ua.edu},
or write to TCGA:
\begin{center}
The Clearinghouse for Genetic Algorithms\\
The University of Alabama\\
Department of Engineering Mechanics\\
P.O. Box 870278\\
Tuscaloosa, Alabama 35487
\end{center}

\subsection*{Acknowledgments}
The authors gratefully acknowledge support provided by NASA under
Grant NGT--50224 and support provided by the 
National Science Foundation under Grant CTS--8451610.
We also thank Hillol Kargupta for donating his tournament selection
implementation.


\bibliography{sga-c}
\end{document}
